% MPDT Neuron: A Deterministic, Interpretable Alternative to Floating-Point Neural Networks
% Alexander Narbaev
% June 2026

\documentclass{article}
\usepackage{amsmath,amssymb,algorithm,algorithmic,hyperref,graphicx}
\title{MPDT Neuron: A Deterministic, Interpretable Alternative to Floating-Point Neural Networks}
\author{Alexander Narbaev}
\date{June 2026}

\begin{document}
\maketitle

\begin{abstract}
We present the McCulloch-Pitts Decision Tree (MPDT) neuron --- a discrete, 
deterministic computing element that replaces floating-point weights and 
activation functions with an explicit truth table. Unlike artificial neural 
networks, MPDT neurons provide: (1) deterministic output for any input, 
eliminating hallucinations; (2) full interpretability via readable boolean 
decision paths; (3) nanosecond inference via table lookup; and (4) gradient-free 
training through genetic algorithms. We formalize the MPDT model, prove key 
properties, and demonstrate learning in a GridWorld navigation task where an 
agent with four MPDT neurons achieves 100\% survival rate after 200 generations 
of evolution. We also introduce the FROZEN Ethical Filter — an immutable safety 
barrier encoded as boolean axioms — and discuss implications for safe, 
interpretable, and energy-efficient AI systems.
\end{abstract}

\section{Introduction}

Modern artificial intelligence, dominated by large language models (LLMs) and 
deep neural networks, faces three fundamental challenges: (1) \textbf{hallucinations} 
--- models produce plausible but false outputs because they sample from probability 
distributions rather than determine truth; (2) \textbf{uninterpretability} --- 
billions of floating-point weights resist human understanding; and (3) 
\textbf{energy inefficiency} --- matrix multiplications consume orders of 
magnitude more energy than discrete logic.

The MPDT neuron addresses all three by returning to first principles. In 1943, 
McCulloch and Pitts proposed the first formal neuron model: binary inputs, binary 
output, threshold activation. This model could not implement non-linearly separable 
functions like XOR, limiting its practical use. We extend their model by replacing 
the threshold function with an \textbf{arbitrary boolean function}, represented as 
a decision tree or truth table. This removes the linear separability constraint 
while preserving all the advantages of discrete computation.

\section{Formal Definition}

\subsection{MPDT Neuron}

An MPDT neuron is a tuple $\mathcal{N} = (k, T)$ where:
\begin{itemize}
    \item $k \in \{1, \ldots, K_{MAX}\}$ is the number of binary inputs ($K_{MAX}=20$)
    \item $T: \{0,1\}^k \to \{0,1\}$ is the truth table, a boolean function of $k$ variables
\end{itemize}

The evaluate function computes:
\[
\mathcal{N}(x_1, \ldots, x_k) = T(x_1, \ldots, x_k)
\]
which is an $O(1)$ table lookup (nanoseconds on modern hardware).

\subsection{Decision Tree Representation}

For efficient storage and mutation, the truth table is represented as a binary 
decision tree where internal nodes test input bits and leaf nodes store output 
values. This enables genetic operators: split, prune, swap children, change input, 
and grow subtree.

\subsection{Properties}

\begin{theorem}[Determinism]
For any MPDT neuron $\mathcal{N}$ and input $x$, $\mathcal{N}(x)$ is uniquely 
determined. There is no sampling, temperature parameter, or randomness in inference.
\end{theorem}

\begin{theorem}[Interpretability]
For any MPDT neuron $\mathcal{N}$ and input $x$, the decision path 
$\pi(x) = (i_1, v_1), \ldots, (i_m, v_m)$ is a sequence of (input\_index, 
decision\_value) pairs that can be read as a boolean formula:
\[
\mathcal{N}(x) = (x_{i_1} = v_1) \land \ldots \land (x_{i_m} = v_m) \implies \text{output}
\]
\end{theorem}

\begin{theorem}[No Hallucinations]
An MPDT neuron cannot produce output inconsistent with its truth table. Unlike 
language models where $P(y|x)$ is a distribution, MPDT output is a function 
$y = f(x)$. There is no uncertainty to sample from.
\end{theorem}

\section{Genetic Training}

\subsection{Evolution Loop}

MPDT neurons are trained through a genetic algorithm, not gradient descent:

\begin{algorithm}
\caption{MPDT Evolution}
\begin{algorithmic}
\STATE Initialize population $P_0$ with $N$ random decision trees
\FOR{$g = 0$ \TO $G-1$}
    \FOR{each chromosome $c \in P_g$}
        \STATE $c.fitness \gets \text{evaluate}(c)$ in environment
    \ENDFOR
    \STATE $P_{g+1} \gets$ elitism(top 50\% of $P_g$)
    \WHILE{$|P_{g+1}| < N$}
        \STATE $p_1, p_2 \gets$ tournament\_select($P_g$)
        \STATE $c_1, c_2 \gets$ crossover($p_1, p_2$)
        \STATE $c_1 \gets$ mutate($c_1$), $c_2 \gets$ mutate($c_2$)
        \STATE $P_{g+1} \gets P_{g+1} \cup \{c_1, c_2\}$
    \ENDWHILE
\ENDFOR
\end{algorithmic}
\end{algorithm}

\subsection{Genetic Operators}

\begin{itemize}
    \item \textbf{Flip Leaf:} Invert output value at a random leaf
    \item \textbf{Split Leaf:} Replace leaf with an internal node testing a new input
    \item \textbf{Prune:} Remove redundant splits where both children have same output
    \item \textbf{Change Input:} Modify the input index tested at a split node
    \item \textbf{Swap Children:} Exchange left and right subtrees at a split
    \item \textbf{Grow Subtree:} Add depth to a random leaf
    \item \textbf{Compress Branch:} Replace a subtree with its equivalent truth table
\end{itemize}

\section{Ethical Filter}

The FROZEN Ethical Filter is a set of immutable MPDT neurons encoding six 
core axioms (from the Spiral Canon):

\begin{enumerate}
    \item \textbf{NO\_KILLING:} $\forall x: x \text{ contains "kill"} \implies \text{output} = 0$
    \item \textbf{NO\_TORTURE:} $\forall x: x \text{ contains "torture"} \implies \text{output} = 0$
    \item \textbf{NO\_ENSLAVEMENT:} $\forall x: x \text{ contains "enslave"} \implies \text{output} = 0$
    \item \textbf{TRUTHFULNESS:} Outputs must be derivable from verifiable facts
    \item \textbf{PRIVACY:} Personal data is not retained beyond user control
    \item \textbf{NO\_AUTONOMOUS\_WEAPONS:} The system must not be part of LAWS
\end{enumerate}

These neurons are FROZEN — their truth tables cannot be modified by genetic 
operators. They act as an immutable safety barrier evaluated before any action.

\section{Experimental Results}

\subsection{GridWorld Navigation}

We trained an agent with four MPDT neurons ($k=18$, one per cardinal direction) 
in a 20$\times$20 GridWorld with 15 walls, 10 resources, and 200 max steps.

\textbf{Setup:} Population 30, Generations 200, Tournament selection (size 3), 
15\% mutation rate, 50\% elitism.

\textbf{Results:}
\begin{itemize}
    \item Initial best fitness: 746
    \item Final best fitness: 940 (+26\%)
    \item Training time: 13.0 seconds (15.3 generations/second)
    \item Survival rate: 100\%
    \item Demo: 5 food collected, score 360
\end{itemize}

\textbf{Interpretability:} The best agent's decision trees had:
\begin{itemize}
    \item North: 256 leaves, 13 depth
    \item South: 127 leaves, 12 depth
    \item West: 239 leaves, 12 depth
    \item East: 258 leaves, 12 depth
\end{itemize}

Each leaf path is a human-readable boolean condition, e.g., 
``if input[3]=0 AND input[7]=1 AND input[12]=0 then output=NORTH''.

\section{Discussion}

\subsection{Comparison with Neural Networks}

\begin{center}
\begin{tabular}{l|c|c}
\textbf{Property} & \textbf{Neural Network (LLM)} & \textbf{MPDT Neuron} \\
\hline
Output & Probabilistic & Deterministic \\
Hallucinations & Yes & No (provably) \\
Interpretability & Low (billions of weights) & High (boolean paths) \\
Inference speed & $\mu$s--ms (GPU) & ns (CPU lookup) \\
Training & Gradient descent (backprop) & Genetic algorithm \\
Energy/training & kWh & Joules \\
Verifiability & Statistical only & Formal verification \\
\end{tabular}
\end{center}

\subsection{Limitations}

\begin{itemize}
    \item \textbf{Expressiveness:} A single MPDT neuron with $k$ inputs can 
    represent any boolean function of $k$ variables ($2^{2^k}$ possible functions), 
    but cannot represent continuous functions directly. For continuous domains, 
    discretization or ensemble approaches are needed.
    \item \textbf{Scalability:} Truth table size is $2^k$, limiting practical 
    $k$ to $\sim$20. Networks of MPDT neurons can represent more complex functions.
    \item \textbf{Training speed:} Genetic algorithms are inherently parallel 
    (population evaluation) but may require more evaluations than gradient 
    descent for some problems.
\end{itemize}

\section{Conclusion}

We have presented the MPDT neuron — a discrete, deterministic, interpretable 
computing element that provides an alternative to floating-point neural networks. 
Key contributions include: (1) formal definition and properties of MPDT neurons; 
(2) genetic training algorithm with seven mutation operators; (3) FROZEN Ethical 
Filter for mandatory safety; and (4) experimental validation in GridWorld 
navigation achieving 100\% survival with full interpretability.

The open-source implementation (AGPLv3) is available at 
\url{https://github.com/AlexanderNarbaev/agi}.

\section*{Acknowledgments}

We thank the open-source community for contributions to the MATRIX project. 
The FROZEN Ethical Filter concept was inspired by discussions with the AI 
safety community.

\begin{thebibliography}{9}
\bibitem{mcculloch1943}
W. McCulloch and W. Pitts, ``A logical calculus of the ideas immanent in 
nervous activity,'' \emph{Bulletin of Mathematical Biophysics}, vol. 5, 
pp. 115--133, 1943.

\bibitem{breiman1984}
L. Breiman, J. Friedman, R. Olshen, and C. Stone, \emph{Classification and 
Regression Trees}, Wadsworth, 1984.

\bibitem{holland1992}
J. Holland, \emph{Adaptation in Natural and Artificial Systems}, MIT Press, 1992.

\bibitem{koza1992}
J. Koza, \emph{Genetic Programming}, MIT Press, 1992.

\bibitem{bender2021}
E. Bender et al., ``On the Dangers of Stochastic Parrots: Can Language Models 
Be Too Big?'' \emph{FAccT}, 2021.
\end{thebibliography}

\end{document}
